Synthesizing Quotient Lenses

### 2018-03-12 papers/QuotientLenses/main.tex
More edits

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}

\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

\newcommand{\QRESize}{\textbf{QS}}
\newcommand{\CanonizerAndSpecSize}{\textbf{CS}}
\newcommand{\LensAndSpecSize}{\textbf{LS}}

@@

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made to data in one format, data in the other must be updated so the
two representations continue to store the same information.
One way to synchronize such data is to use \emph{lenses}\ksf{include citations?}, which
are programs that simultaneously define pairs of functions for translating
the data 
from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses are usually
defined using domain-specific bidirectional languages that 
ensure the specified lenses satisfy various \emph{lens laws} that
guarantee how the \emph{put} and \emph{get} functions interact.

@@

The second drawback with writing quotient lenses in Boomerang is that
even after specifying the four required data formats, the programmer must
still define a lens for converting between the canonical
representations.   Experience shows that writing such lenses can be
tedious 
(consider having to rearrange the order of data items by recursively
using operators to \swap adjacent fields) 
and that satisfying the type checker can be difficult because 
many seemingly useful lenses contain ambiguities disallowed by the 
Boomerang type checker~\cite{optician}.

@@

\section{Quotient Regular Expressions}

\BCP{What became of my suggestion that we start with just ``normalize'' and
  then, later, introduce the other, more specific (and pragmatic, and
  incomplete) ones?}

A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.
\label{QRE}
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
Q := \; &R \sep \project{R}{s} \sep \squash{Q_1}{Q_2}{f} \sep
\perm{Q_1, \ldots, Q_n}{Q} \\
& | \; \normalize{R}{R'}{f} \sep Q_1 \semicolon Q_2 \sep Q_1 \cdot Q_2 \sep (Q_1
\sep Q_2) \sep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $Q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(Q)$ (the ``whole'' of $Q$),
  \item an equivalence relation $\eqrel{Q}$ on $\mathcal{L}(W(Q))$,
  \item a regular expression $K(Q)$ (the ``kernel'' of $Q$)
  such that $\mathcal{L}(K(Q))$ forms a complete set of representatives for
  $\eqrel{Q}$, and
  \item a ``canonizing'' function\bcp{Why is it called ``canonizer'' rather
  than ``canonize''??}\sam{because \canonizer(Q) is a function}\bcp{My ear
  consistently finds it quite awkward.  I'd be happier with
  ``canonize'' (which describes an action).}
  $\canonizer(Q):\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$ which
  given any $w \in \mathcal{L}(W(Q))$, computes $\canonizer(Q)(w)$ as the
  unique $k$ in $\mathcal{L}(K(Q))$ such that $k$ is equivalent to $w$ mod
  $\eqrel{Q}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(Q)$ and $K(Q)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q}{Q_1}{f} & W(Q) \sep W(Q_1) & K(Q_1) \\
\normalize{R}{R'}{f} & R & R' \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$. This is because the \textit{squash} combinator
takes the whole language $W(Q_1)$ of $Q_1$, the whole language $W(Q_2)$ of $Q_2$
and a function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonizer(Q_2)$.

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonizer(R) &=& id_{\mathcal{L}(R)} \\
\canonizer(\project{R}{s})(w) &=& s\\
\canonizer(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonizer(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonizer(Q_2)(w) & \text{otherwise}
\end{cases}\\
\canonizer(\normalize{R}{R'}{f}) &=& f\\
\canonizer(Q_1 \; ; \; Q_2) &=& \canonizer(Q_2) \circ \canonizer(Q_1)\\
\canonizer(Q_1 \cdot Q_2) &=& \canonizer(Q_1) \cdot \canonizer(Q_2)\\
\canonizer(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonizer(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonizer(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\canonizer(Q^*) &=& \canonizer(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonizer(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonizer(Q_1)(r_1)) \cdot (\canonizer(Q)(s_1)) \cdot \ldots \cdot
(\canonizer(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \textit{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonizer(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonizer(Q_1)(r_1)) \cdot (\canonizer(Q)(s_1)) \cdot \ldots \cdot
(\canonizer(Q_n)(r_n))
\end{align*}
\noindent which, as we have explained, is defined so that the strings that match
$Q_i, Q$ are placed in the canonical permutation which is the identity
permutation before $\canonizer(Q_i)$ and $\canonizer(Q)$ are applied
respectively.

@@

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_R \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Well Formed QREs}
\label{subsec:well-formed-qres}
\subsubsection{Completeness of {\em normalize}}
Before we present the full inference rules, for QREs, we discuss the {\em
normalize} QRE which is special in two main ways, the first of which has to do
with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

We do this firstly because the {\em normalize} combinator allows one to express
equivalence relations that are either difficult to define, or that cannot be
defined using the other combinators. Secondly, the {\em normalize} canonizer
enjoys the property that each of the other combinators $Q$ is semantically
equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some surjective, idempotent
function $f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$; this is the
second property that distinguishes the {\em normalize} combinator from the other
QREs. The {\em normalize} combinator hence gives a sufficient condition for
potential QREs to be ``well-behaved'', particularly with respect to the
strategy for synthesizing quotient lenses which we shall introduce in
Section~\ref{synth}. This is the
\subsubsection{Ambiguity}
Another issue which we address before presenting the full set of inference
rules for QREs, we discuss the issue of {\em ambiguity}, which appears in the
premises of the inference rules for the regular combinators. If $Q_1$ and $Q_2$
are QREs, then when applying the regular combinators to QREs, we require that
if a string $s$ matches any of the regular expressions $W(Q_1) \cdot W(Q_2)$,
\quad $W(Q_1) \sep W(Q_2)$, \quad $W(Q_1)^*$, \quad $K(Q_1) \cdot K(Q_2)$,
\quad $K(Q_1) \sep K(Q_2)$, \quad $K(Q_1)^*$, then $s$ matches that regular
expression in only one way.

@@

However, we also have a secondary objective, which is to support synthesis of
lenses: When provided with QREs describing the source and target languages, we
would like to be able to generate quotient lenses automatically. One way to do
that is to generate canonizers $\canonizer(Q_1)$ and $\canonizer(Q_2)$ from QREs
$Q_1$ and $Q_2$ and then to synthesize bijective lenses between the kernel
languages for $Q_1$ and $Q_2$.

@@

$$\ell := \mathit{id} \; (R) \sep \const(s_1, s_2) \sep  \swap(\ell_1, \ell_2)
\sep \ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \; | \;
\ell_1 \; ; \;  \ell_2,$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The $\mathit{\const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the
backward direction, while $\mathit{id}(R)$ applies the identity function to the
source and target  in $\mathcal{L}(R)$ in both directions. The composition lens
$\ell_1 \; ; \; \ell_2$ applies $\ell_1$ followed by then $\ell_2$ to the
source in the forward direction, and applies $\ell_2$ followed by $\ell_1$ to
the target in the backward direction. The lens $\ell_1 \cdot \ell_2$ first
splits the string $s$ into $s_1$ and $s_2$, applies $\ell_1$ and $\ell_2$ to
$s_1$ and $s_2$ to get $t_1$ and $t_2$ respectively, then concatenates $t_1$
and $t_2$ and returns $t_1 \cdot t_2$ as the final result in the forward
direction. $\ell_1 \cdot \ell_2$ operates similarly in the backward direction,
but with $s, s_1$ and $s_2$ substituted for $t, t_1$ and $t_2$. The
$\mathit{\swap} \; (\ell_1, \ell_2)$ lens operates like $\ell_1 \cdot \ell_2$,
except that it swaps $t_1$ and $t_2$ before concatenating the two for a final
result of $t_2 \cdot t_1$ in the forward direction. In the backward direction,
$\mathit{\swap}(\ell_1, \ell_2)$ first undoes the \swap, then behaves like
$cdot$. The $\ell_1 \; | \; \ell_2$ lens chooses to apply $\ell_1$ or $\ell_2$
depending on whether the source (resp. target) data is matched by $\ell_1$ or
$\ell_2$ in the forward (resp. backward direction). The $\ell^*$ lens splits the
string $s$ into strings $s_1, \ldots, s_n$, applies $\ell_1$ to each $s_i$ to
get $t_i$, and then concatenates each of the $t_i$'s for a final result of $t_1
\cdot \ldots \cdot t_n$ in the forward direction. The iteration lens $\ell^*$
operates similarly in the backward direction, but with $s, s_i$ substituted for
$t, t_i$.

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

\newcommand{\QRESize}{\textbf{QS}}
\newcommand{\canonizeAndSpecSize}{\textbf{CS}}
\newcommand{\LensAndSpecSize}{\textbf{LS}}

@@

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made to data in one format, data in the other must be updated so the
two representations continue to store the same information.
One way to synchronize such data is to use \emph{lenses}~\cite{Lenses}, which
are programs that simultaneously define pairs of functions for translating
the data from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses are usually
defined using domain-specific bidirectional languages that 
ensure the specified lenses satisfy various \emph{lens laws} that
guarantee how the \emph{put} and \emph{get} functions interact.

@@

The second drawback with writing quotient lenses in Boomerang is that
even after specifying the four required data formats, the programmer must
still define a lens for converting between the canonical
representations.   Experience shows that writing such lenses can be
tedious (consider having to rearrange the order of data items by recursively
using operators to \kw{swap} adjacent fields)  and that satisfying the type
checker can be difficult because many seemingly useful lenses contain
ambiguities disallowed by the Boomerang type checker~\cite{optician}.

@@

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}, but first we discuss the \kw{normalize} combinator which
enjoys some special properties.

\subsection{The \kw{normalize} Combinator}
The \kw{normalize} combinator is special in two main ways, the first of
which has to do with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

We do this firstly because the {\em normalize} combinator allows one to express
equivalence relations that are either difficult to define, or that cannot be
defined using the other combinators. Secondly, the {\em normalize} canonizer
enjoys the property that each of the other combinators $Q$ is semantically
equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some surjective, idempotent
function $f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$; this is the
second property that distinguishes the {\em normalize} combinator from the other
QREs. The {\em normalize} combinator hence gives a sufficient condition for
potential QREs to be ``well-behaved'', particularly with respect to the
strategy for synthesizing quotient lenses which we shall introduce in
Section~\ref{synth}.

Having discussed the special properties of the \kw{normalize} combinator, we now
give the syntax and semantics of the other QRE combinators.
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
Q := \; &\id(R) \sep \project{R}{s} \sep \squash{Q_1}{Q_2}{f} \sep
\perm{Q_1, \ldots, Q_n}{Q} \\
& | \; \normalize{R}{R'}{f} \sep Q_1 \semicolon Q_2 \sep Q_1 \cdot Q_2 \sep (Q_1
\sep Q_2) \sep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $Q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(Q)$ (the ``whole'' of $Q$),
  \item an equivalence relation $\eqrel{Q}$ on $\mathcal{L}(W(Q))$,
  \item a regular expression $K(Q)$ (the ``kernel'' of $Q$)
  such that $\mathcal{L}(K(Q))$ forms a complete set of representatives for
  $\eqrel{Q}$, and
  \item a ``canonizing'' function $\canonize(Q):\mathcal{L}(W(Q))
  \longrightarrow \mathcal{L}(K(Q))$ which given any $w \in \mathcal{L}(W(Q))$,
  computes $\canonize(Q)(w)$ as the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k$ is equivalent to $w$ mod $\eqrel{Q}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(Q)$ and $K(Q)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q}{Q_1}{f} & W(Q) \sep W(Q_1) & K(Q_1) \\
\normalize{R}{R'}{f} & R & R' \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$. This is because the \textit{squash} combinator
takes the whole language $W(Q_1)$ of $Q_1$, the whole language $W(Q_2)$ of $Q_2$
and a function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\project{R}{s})(w) &=& s\\
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\
\canonize(\normalize{R}{R'}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \textit{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\noindent which, as we have explained, is defined so that the strings that match
$Q_i, Q$ are placed in the canonical permutation which is the identity
permutation before $\canonize(Q_i)$ and $\canonize(Q)$ are applied
respectively.

@@

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well Formed QREs}
\label{subsec:well-formed-qres}
An issue that we address before presenting the full set of inference
rules for QREs is the issue of {\em ambiguity}, which appears in the
premises of the inference rules for the regular combinators. If $Q_1$ and $Q_2$
are QREs, then when applying the regular combinators to QREs, we require that
if a string $s$ matches any of the regular expressions $W(Q_1) \cdot W(Q_2)$,
\quad $W(Q_1) \sep W(Q_2)$, \quad $W(Q_1)^*$, \quad $K(Q_1) \cdot K(Q_2)$,
\quad $K(Q_1) \sep K(Q_2)$, \quad $K(Q_1)^*$, then $s$ matches that regular
expression in only one way.

@@

However, we also have a secondary objective, which is to support synthesis of
lenses: When provided with QREs describing the source and target languages, we
would like to be able to generate quotient lenses automatically. One way to do
that is to generate canonizers $\canonize(Q_1)$ and $\canonize(Q_2)$ from QREs
$Q_1$ and $Q_2$ and then to synthesize bijective lenses between the kernel
languages for $Q_1$ and $Q_2$.

@@

$$\ell := \id(R) \sep \const(s_1, s_2) \sep  \swap(\ell_1, \ell_2)
\sep \ell_1 \cdot \ell_2 \; |  \; (\ell_1 \sep \ell_2) \sep \ell^* \; | \;
\ell_1 \; ; \;  \ell_2,$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The $\mathit{\const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the
backward direction, while $\mathit{id}(R)$ applies the identity function to the
source and target  in $\mathcal{L}(R)$ in both directions. The composition lens
$\ell_1 \; ; \; \ell_2$ applies $\ell_1$ followed by then $\ell_2$ to the
source in the forward direction, and applies $\ell_2$ followed by $\ell_1$ to
the target in the backward direction. The lens $\ell_1 \cdot \ell_2$ first
splits the string $s$ into $s_1$ and $s_2$, applies $\ell_1$ and $\ell_2$ to
$s_1$ and $s_2$ to get $t_1$ and $t_2$ respectively, then concatenates $t_1$
and $t_2$ and returns $t_1 \cdot t_2$ as the final result in the forward
direction. $\ell_1 \cdot \ell_2$ operates similarly in the backward direction,
but with $s, s_1$ and $s_2$ substituted for $t, t_1$ and $t_2$. The
$\mathit{\swap} \; (\ell_1, \ell_2)$ lens operates like $\ell_1 \cdot \ell_2$,
except that it swaps $t_1$ and $t_2$ before concatenating the two for a final
result of $t_2 \cdot t_1$ in the forward direction. In the backward direction,
$\mathit{\swap}(\ell_1, \ell_2)$ first undoes the \swap, then behaves like
the concatenation lens. The $\ell_1 \; | \; \ell_2$ lens chooses to apply
$\ell_1$ or $\ell_2$ depending on whether the source (resp. target) data is
matched by $\ell_1$ or $\ell_2$ in the forward (resp. backward direction). The
$\ell^*$ lens splits the string $s$ into strings $s_1, \ldots, s_n$, applies
$\ell_1$ to each $s_i$ to get $t_i$, and then concatenates each of the $t_i$'s
for a final result of $t_1 \cdot \ldots \cdot t_n$ in the forward direction.
The iteration lens $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.
